An Empirical Analysis of the Role of Energy in Economic Growth

### 2014-12-27 Articles/Article_EnerEcon_2014/EconGrowth3.Rnw
Moved lit review to Introduction.

Many who accept that energy use is closely correlated with
economic performance, regardless of the direction of causality,
express concern regarding the negative effects of 
rising energy prices and fossil fuel depletion 
on current and future economic growth.
\citet{Benes:2012tm}, incorporating views on both
geological physical constraints and 
technological solutions to oil production, 
suggest a near doubling of the real price of over the next decade.  
\citet{Kumhof:2012ua} point out that 
significant impacts on GDP growth could occur
for larger oil supply shocks, 
assuming production functions where oil plays a prominent role 
and reduced possibilities for substitution away from oil.  
\citet{Nel:2010fv} argue for the 
``\dots{}vital importance of energy security\dots'' 
on the basis of an economic growth model with explicit 
energy-based formulation.  
\citet{Hamilton:2013vc} argues that 
``[c]oping with a final peak in world oil production\dots%
[will] have much in common with previous historical episodes that resulted from 
temporary geopolitical conflict, 
being associated with significant declines in employment and output. 
If the future decades look like the last 5 years, we are in for a rough time.''

@@

As shown in Equation~\ref{eq:CD} above, mainstream economic theory does not 
include an explicit term for the role of energy in economic growth. 
However, \citet{Bashmakov:2007ek}, \citet{Hamilton:2009ui}, \citet{Murphy:2011jh}, 
\citet{Benes:2012tm}, \citet{Kumhof:2012ua}, \citet{Ayres:2013aa}, and others indicate that the 
importance of energy for economic growth may far outweigh its cost share in the 
economy. 
That mainstream economic growth theory ignores energy is, therefore, 
cause for concern, and inquiry into the proper approach for including energy 
as a factor of production in economic growth modeling is both relevant and important.

Furthermore, the \emph{cost share theorem} from mainstream economics states that 
output elasticities 
$\alpha$, $\beta$, and $\gamma$ 
(if augmenting the production function by energy)
can be calibrated from national accounts data as the 
cost shares of capital, labor, and energy, respectively.
However, the cost share theorem applies only under conditions of perfect competition
in the markets for the factors of production.
Although no market ever exhibits perfect competition, 
in our opinion, energy markets are \emph{least} likely to be ``perfect.''
Many governments have parastatal or full control 
of electricity generation and/or fuel markets.  
Other governments subsidize or incentivize energy markets.  
And, there are always health and environmental externalities 
that are not priced into fossil fuel energy markets.  
If energy markets are farther from ``perfect'' 
than capital~($K$) and/or labor~($L$) markets, 
then energy is the factor of production least likely to have its
output elasticity~($\gamma$) be well-represented 
by cost share as reported in national accounts.
Thus, when augmenting production functions by energy, 
much care should be taken in estimating 
the relative importance of the factors of production 
($\alpha$, $\beta$, and $\gamma$).

\citet{Kummel:2010vz} showed that the cost share theorem applies 
only when economic growth is unconstrained.
Real-world constraints such as the degree of automation and
the degree of utilization of capital stock 
result in shadow prices that move the values 
of $\alpha$, $\beta$, and $\\gamma$ away from 
their cost shares in the economy.\citep[Equations~12 and~13]{Kummel:2010vz}
Thus, for this study, 
we reject the cost share theorem and
we have no \emph{a-priori} expectation that output elasticities 
estimated from historical output data
will match cost shares.

Some energy economists augment economic growth models by including energy as an 
additional (or the only) factor of production. 
However, there is little agreement on how this should be done. 
Neither the mathematical form of the energy-augmented 
production function nor the quantification of energy%
  \footnote{
  Various energy quantifications have been used in economic growth modeling: 
  thermal energy (primary and final), 
  exergy (primary and final), and useful work.
  }
is consistent in the literature.

**** Suggest moving content of the literature review section up to this point. --MKH ****

**** Suggest deleting the next sentences:
Furthermore, there has been little empirical work assessing the predictive ability 
of various energy-augmented production functions for different types of economies. 
And, very few studies of the importance of energy in economic growth include 
rigorous statistical assessments of energy-related fitting parameters. 
End of deleted section --MKH ****

@@

The remainder of this paper is organized as follows. 
Section~\ref{sec:lit_review} provides a necessarily brief overview of relevant literature.
Section~\ref{sec:Coordinates} discusses the coordinates of analysis for the present study. 
Section~\ref{sec:Sources_of_Data} details sources of data and calculations to 
generate time series for economic output ($y$) and factors of production ($k$, $l$, and $q$). 
The procedures for and results from parameter estimation for the various production functions 
are discussed in Section~\ref{sec:Parameter_Estimation}. 
Section~\ref{sec:Discussion_of_Results} discusses implications of the parameter estimation 
results, and Section~\ref{sec:Conclusion} concludes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Literature Review} 
\label{sec:lit_review}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

**** Merge the literature review into the Introduction? 
We're already doing some literature review in the Introduction. --MKH ****

\citet[p.~26]{Stern:2011bq} provides an overview of approaches to including energy 
in ecological economic and resource economic models of growth.  
\citeauthor{Stern:2011bq} points out that 
energy is seen as the \emph{central driver of growth} 
in ecological economic (including biophysical economic) models
and that energy \emph{is included in} resource economic models, 
but \emph{remains isolated} within the sub-discipline.  
To break through this theoretical impasse, 
Stern proposes a synthesized Constant Elasticity of Substitution (CES) 
production function 
where the constraints or abundance of energy 
determines whether to include energy explicitly 
as a factor of production.  
On the basis of this model, 
\citet[p. 45]{Stern:2011bq} points out that 
energy availability plays
``\dots{}a key role in enabling growth,''
despite evidence of decreasing energy consumed per unit of output.  
He concludes, based on theory and time-series results, 
that including energy in the production function is warranted, 
and that further clarification is needed 
on the role of energy in economic growth.

Unsatisfactorily, the choice of a production function 
and the parameters used to populate these functions 
affects results significantly.
This situation leaves far too much latitude 
for misuse of model specifications 
and selection of model parameters 
to support certain theories (rather than empirically testing the theories
and parameters \emph{before} they are employed).  
For example, \citet{Lecca:2011kr} pointed out that 
the modellers' choice for the structure of a production function 
reflects his or her assumptions about
the elasticities among inputs, and 
``\dots the point at which energy enters the [nested CES] production function 
has a significant impact on CGE [computable general equilibrium] model results.''
\citet{Duffy:2000ty} rejected the Cobb-Douglas production function altogether, 
finding that capital and labour are more substitutable 
in richer countries and less substitutable in poorer countries 
than Cobb-Douglas implies.  
However, \citet{Miller:2008vd} points to the 
``\dots reasonably accurate long-term economic forecasts,'' 
achieved with the use of Cobb-Douglas production functions in economic models.  
\citet{Shen:2013vp} provide a recent example 
wherein empirical analysis on the estimated substitution elasticities 
is performed before choosing any particular 
nested CES production function structure. 
As a result, \citeauthor{Shen:2013vp} propose an atypical CES nesting structure 
for the Chinese economy. **** More detail here about nesting? --MKH ****  
\citet{Saunders:2008aa} evaluated several production and cost functions
(including the Cobb-Douglas, CES, and LINEX functions studied herein)
to assess whether they are capable of reproducing 
rebound, backfire, and super-conservation effects.~\citep{Sorrell:2007um}
\citeauthor{Saunders:2008aa} concludes that only the CES function is a candidate for 
rebound analysis.
The Cobb-Douglas function was found to exhibit only backfire.
The LINEX function was found to 
``not satisfy the standard concavity (diminishing marginal productivity) 
conditions.''~\citet[p.~2195]{Saunders:2008aa}
Both the choice of production function structure 
and the value of the model parameters 
drive model results, and empirical and statistical testing 
of such approaches is prudent, but rarely conducted.

These are not idle academic debates. 
Macroeconomic growth models are used often to inform economic policy,
and the paths of economies worldwide are determined by
such policy choices. 
Therefore, we agree with those who have serious concerns 
about the reliability of the results from economic growth models \citep{Lecca:2011kr}. 
We understand, however, 
that macroeconomic policymaking is needed 
and is increasingly informed by integrated energy-economic models, 
in a world where energy constraints are possible 
\citep{Beckman:2011gb, Benes:2012tm, Nel:2010fv, Kummel:2010vz}.  
We support the call from \citet{Shen:2013vp} that both
the structure of the production function 
and the parameters that populate it 
need empirical testing. 
Building on the argument of \citet{Stern:2011bq} 
that energy may or may not be required in production functions 
depending on its abundance or scarcity, 
we will further argue in this paper that different types of countries 
may need different factors of production to explain their respective growth paths.  
This further leaves open the possibility for 
different explanations of growth in various countries
at different stages of their development, 
a central point of long-wave economic growth modeling 
\citep{Froling:2009vo, Jones:2001wn, Galor:2000up}. 
**** The previous sentences sound a lot like we'll show that some groups of countries
have similar charactersitics. 
That's what we thought we might find, 
but our new conclusion is that one must treat each country 
individually. 
We should consider revising the few sentences. --MKH ****

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion of Results}
\label{sec:Discussion_of_Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We find it beneficial to discuss results through a series of questions, 
answers to which are illuminated by the techniques and results presented above.

@@

For the Cobb-Douglas and CES production functions, 
augmentation by energy improves goodness of fit when $\gamma > 0$.%
  \footnote{
  The LINEX function is unhelpful in answering the question 
  ``Is it important to include energy in production functions?'' 
  because LINEX presupposes that energy is the only factor of production.
  }

First, we look at the results of the Cobb-Douglas functions 
in Figures~\ref{fig:CD_lambda_theta_graph} and~\ref{fig:CD_Ternary_Scatter_Graph}. 
Visually, $\gamma > 0$ is shown in the right column 
of Figure~\ref{fig:CD_Ternary_Scatter_Graph} 
as data points that are above the horizontal base of the ternary plots. 
We note that the list of countries for which $\gamma \gg 0$ includes 
Tanzania, South Africa, Saudi Arabia, and the United Kingdom. 
Especially for Tanzania, but also for Saudi Arabia and the United Kingdom, 
it appears certain that including energy improves goodness of fit: 
both the fit to historical data (crosshairs) 
and all resample data points (black dots) 
are above the base of the triangle in the energy columns of Figure~\ref{fig:CD_Ternary_Scatter_Graph}. 
However, for South Africa, significant uncertainty in $\gamma$, 
as shown by the large vertical spread of resample data points, 
leads us to exercise caution in concluding that adding energy to the production function is helpful
for improving fit to historical data. 
A robust result 
($\gamma$ equal or close to zero and very little vertical spread of the resample points) 
indicates that including energy does not help to explain economic growth in Zambia,
while for United States, Japan, China, and especially Iran, 
it is less certain that augmenting with energy is unhelpful 
(larger vertical spread of resample points).

@@

Based on this analysis of the empirical data, 
we find no grounds to argue for augmenting the production function 
with energy for all economies. 
Our results support \citeapos{Stern:2011bq} conclusion that energy 
may or may not be required in production functions,
depending on its abundance or scarcity. 
But, one cannot know \emph{a-priori} the economies 
for which augmenting by energy will improve goodness of fit. 
Statistical testing is clearly warranted.

@@

Some developed economies (US) benefit little from including energy as a factor of production, 
while some others (UK) benefit in a minor way. 
For other developed economies (JP), the results of CD and CES models point in opposite directions.  
For NICs the results are mixed; 
China does not convincingly benefit from energy-augmented production functions, 
and results are uncertain for South Africa (ZA) due to lack of precision in determining $\gamma$.  
In one OPEC country (SA), adding energy clearly helps to explain economic growth, 
while in another (IR), the benefit of energy augmentation is small and uncertain. 
For LDCs there is a either a strong indication that energy-augmented production functions 
are beneficial as an explanation for growth (TZ) or string indication otherwise (ZM).

In summary, we observe that World Bank country categories (Table~\ref{tab:Economy_Examples}) 
do not predict whether energy augmentation will be beneficial. 
In fact, 
there is as much variation of the role of energy in economic growth 
within World Bank country categories
as between World Bank country categories. 
Thus, we conclude that the World Bank’s country categories are not helpful 
in explaining the role of energy in economic growth. 
Rather, a nuanced approach is necessary wherein the energy situation 
in each economy is assessed independently.

@@

The interpretation of elasticities of substitution 
among factors of production ($k$, $l$, and $e$) is challenging,
because a version of each elasticity of substitution appears three or four times 
among the various nests. 
For example, an elasticity of substitution between capital~($k$) and labor~($l$) 
appears in the following nests: 
($\underline{kl}$), 
($\underline{kl}$)$e$, 
($\underline{l}e$)$\underline{k}$, and 
($e\underline{k}$)$\underline{l}$. 
The following tables assist interpretation: 
C (c) indicates strong (weak) complementarity, 
S (s) indicates strong (weak) substitutability, 
$\sim\sim\sim$ indicates large uncertainty, 
and $\sim$ indicates small uncertainty.

@@

Regarding Martin's question about length, 
I've seen other papers in \emph{Energy Economics} that are 51 printed journal pages. (Saunders, 2008)
I expect we are fine on length. --MKH ****

@@

**** For the time being, I've added two of these tables.
We can add the remaining seven tables easily, either here or in an appendix. --MKH ****

@@

That so many estimates of elasticity of substitution exhibit large uncertainty, 
often spanning the range from perfect complements ($\sigma, \sigma_1 = 0$) 
to perfect substitutes ($\sigma, \sigma_1 = 1$), 
is a troubling characteristic of the estimated CES models. 
For some countries (IR, ZM), 
estimating any elasticity of substitution at all is problematic and fraught with peril. 
Note that this result is unlikely to be caused by poor-quality data alone. 
Presumably the best data in the world (US, UK, JP) exhibits large uncertainty. 
(See US and UK rows in Figures~\ref{fig:CD_Ternary_Scatter_Graph}  
and~\ref{fig:CES_sigma1_delta1_graph}--\ref{fig:CES_ternary_graph}.)

@@

Some of the uncertainty in parameter estimates may be explained by structural change in the economies. 
For example, 
**** Martin: can you select one of the highly uncertain economies and discuss it? 
Perhaps Iran or Zambia? ****. 
To account for the large uncertainty, 
it may be useful to perform piecewise analyses of smaller time periods. 
For the present study, doing piecewise analysis of, say, 
Iran **** Use the economy that Martin discusses above **** 
would mean that the CES model is estimated on a 10-year period. 
Considering that there are six fitted parameters in the CES model 
($\gamma$, $\lambda$, $\delta$, $\delta_1$, $\rho$, and $\rho_1$), 
ten years of historical data leaves precious few (only 4) degrees of freedom for the fitting process.

Based on the results and discussions above 
regarding the relative importance of and 
the elasticity of substitution among factors of production, 
we conclude that use of the CES model for energy analysis 
should be undertaken with utmost caution. 
Indeed, empirical testing on the relative importance 
of production factors and elasticities of substitution is highly recommended. 
Sensitivity analysis is essential. 
Our results support \citeapos{Lecca:2011kr} conclusion 
for the need to econometrically specify production function relationships and 
to perform sensitivity analysis on nesting choices in economies to enhance the reliability of results.

@@

Our assessment: None of the production functions is perfect. 
Probably CD is best compromise among ease of estimating, good (but not best) fit to historical data, and
small uncertainties in fitted parameters. 
If precise fitting to historical data is paramount (and you're unconcerned about uncertainties
in fitted coefficients), CES is the best.

Caution: none of the production functions analyzed herein allows for complex energy-economy phenomena
like rebound effect, backfire, and super-conservation.

%++++++++++++++++++++++++++++++
\subsection{Implications of resampling results } 
\label{sec:implications}
%++++++++++++++++++++++++++++++

>>>>

Many who accept that energy use is closely correlated with
economic performance, regardless of the direction of causality,
express concern regarding the negative effects of 
rising energy prices and fossil fuel depletion 
on current and future economic growth.
\citet{Benes:2012tm}, incorporating views on both
geological physical constraints and 
technological solutions to oil production, 
suggest a near doubling of the real price of over the next decade.  
**** Should we re-think this discussion in the light of the shale oil boom and the 
recent drop in oil prices? --MKH ****
\citet{Kumhof:2012ua} point out that 
significant impacts on GDP growth could occur
for larger oil supply shocks, 
assuming production functions where oil plays a prominent role 
and reduced possibilities for substitution away from oil.  
\citet{Nel:2010fv} argue for the 
``\dots{}vital importance of energy security\dots'' 
on the basis of an economic growth model with explicit 
energy-based formulation.  
\citet{Hamilton:2013vc} argues that 
``[c]oping with a final peak in world oil production\dots%
[will] have much in common with previous historical episodes that resulted from 
temporary geopolitical conflict, 
being associated with significant declines in employment and output. 
If the future decades look like the last 5 years, we are in for a rough time.''

@@

As shown in Equation~\ref{eq:CD} above, mainstream economic theory does not 
include an explicit term for the role of energy in economic growth. 
However, \citet{Bashmakov:2007ek}, \citet{Hamilton:2009ui}, \citet{Murphy:2011jh}, 
\citet{Benes:2012tm}, \citet{Kumhof:2012ua}, \cite{Ayres:2010ug}, 
\citet{Ayres:2013aa}, and others indicate that the 
importance of energy for economic growth may far outweigh its cost share in the 
economy. 
That mainstream economic growth theory ignores energy is, therefore, 
cause for concern, and inquiry into the proper approach for including energy 
as a factor of production in economic growth modeling is both relevant and important.

Some energy economists augment economic growth models by including energy as an 
additional (or the only) factor of production. 
However, there is little agreement on how this should be done. 
Neither the mathematical form of the energy-augmented 
production function nor the quantification of energy%
  \footnote{
  Various energy quantifications have been used in economic growth modeling: 
  thermal energy (primary and final), 
  exergy (primary and final), and useful work.
  }
is consistent in the literature. 
\citet[p.~26]{Stern:2011bq} provides an overview of approaches to including energy 
in ecological economic and resource economic models of growth.  
\citeauthor{Stern:2011bq} points out that 
energy is seen as the \emph{central driver of growth} 
in ecological economic (including biophysical economic) models
and that energy \emph{is included in} resource economic models, 
but \emph{remains isolated} within the sub-discipline.  
To break through this theoretical impasse, 
Stern proposes a synthesized Constant Elasticity of Substitution (CES) 
production function 
where the constraints or abundance of energy 
determines whether to include energy explicitly 
as a factor of production.  
On the basis of this model, 
\citet[p. 45]{Stern:2011bq} points out that 
energy availability plays
``\dots{}a key role in enabling growth,''
despite evidence of decreasing energy consumed per unit of output.  
He concludes, based on theory and time-series results, 
that including energy in the production function is warranted, 
and that further clarification is needed 
on the role of energy in economic growth.

Unsatisfactorily, both the choice of a production function 
and the parameters used to populate these functions 
affect results significantly.
This situation leaves far too much latitude 
for misuse of model specifications 
and selection of model parameters 
to support certain theories (rather than empirically testing the theories
and parameters \emph{before} they are employed).  
For example, \citet{Lecca:2011kr} pointed out that 
the modellers' choice for the structure of a production function 
reflects his or her assumptions about
the elasticities among inputs, and 
``\dots the point at which energy enters the [nested CES] production function 
has a significant impact on CGE [computable general equilibrium] model results.''
\citet{Duffy:2000ty} rejected the Cobb-Douglas production function altogether, 
finding that capital and labour are more substitutable 
in richer countries and less substitutable in poorer countries 
than Cobb-Douglas implies.  
However, \citet{Miller:2008vd} points to the 
``\dots reasonably accurate long-term economic forecasts,'' 
achieved with the use of Cobb-Douglas production functions in economic models.  
\citet{Shen:2013vp} provide a recent example 
wherein empirical analysis on the estimated substitution elasticities 
is performed before choosing any particular 
nested CES production function structure. 
As a result, \citeauthor{Shen:2013vp} propose an atypical CES nesting structure 
for the Chinese economy. **** More detail here about nesting? --MKH ****  
\citet{Saunders:2008aa} evaluated several production and cost functions
(including the Cobb-Douglas, CES, and LINEX functions studied herein)
to assess whether they are capable of reproducing 
rebound and backfire effects.~\citep{Sorrell:2007um}
\citeauthor{Saunders:2008aa} concludes that,
of the production functions evaluated in the present study,
only the CES function is a candidate for 
rebound analysis.
The Cobb-Douglas function was found to exhibit only backfire.
The LINEX function was found to 
``not satisfy the standard concavity (diminishing marginal productivity) 
conditions.''~\citet[p.~2195]{Saunders:2008aa}

In addition to model structure, 
the way in which model parameters are determined is also subject to debate.
The \emph{cost share theorem} from mainstream economics states that 
output elasticities 
$\alpha$, $\beta$, and $\gamma$ 
(if augmenting the production function by energy)
can be calibrated from national accounts data as the 
cost shares of capital, labor, and energy, respectively.
However, the cost share theorem applies only under conditions of perfect competition
in the markets for the factors of production.
Although no market ever exhibits perfect competition, 
in our opinion, energy markets are \emph{least} likely to be ``perfect.''
Many governments have parastatal or full control 
of electricity generation and/or fuel markets.  
Other governments subsidize or incentivize energy markets.  
And, there are always health and environmental externalities 
that are not priced into fossil fuel energy markets.  
If energy markets are farther from ``perfect'' 
than capital and/or labor markets, 
then energy is the factor of production least likely to have its
output elasticity~($\gamma$) be well-represented 
by cost share as reported in national accounts.
Thus, when augmenting production functions by energy, 
much care should be taken in estimating 
the relative importance of the factors of production 
($\alpha$, $\beta$, and $\gamma$). 
\citet{Kummel:2010vz} showed that the cost share theorem applies 
only when economic growth is unconstrained.
Real-world constraints such as the degree of automation and
the degree of utilization of capital stock 
result in shadow prices that move the values 
of $\alpha$, $\beta$, and $\gamma$ away from 
their cost shares in the economy.~\citep[Equations~12 and~13]{Kummel:2010vz}
Thus, for this study, 
we reject the cost share theorem and
we have no \emph{a-priori} expectation that output elasticities 
estimated from historical output data
will match cost shares.
Both the choice of production function structure 
and the values of the model parameters 
drive model results, and empirical and statistical testing 
of such approaches is prudent, but rarely conducted.

These are not idle academic debates. 
Macroeconomic growth models are used often to inform economic policy,
and the paths of economies worldwide are determined by
such policy choices. 
Therefore, we agree with those who express serious concerns 
about the reliability of the results from economic growth models.~\citep{Lecca:2011kr} 
We understand, however, 
that macroeconomic policymaking is needed 
and is increasingly informed by integrated energy-economic models, 
in a world where energy constraints are possible.~\citep{Beckman:2011gb, 
Benes:2012tm, Nel:2010fv, Kummel:2010vz}
We support the call from \citet{Shen:2013vp} that both
the structure of the production function 
and the parameters that populate it 
need empirical testing. 
Building on the argument of \citet{Stern:2011bq} 
that energy may or may not be required in production functions 
depending on its abundance or scarcity, 
we will further test whether different types of countries 
may need different factors of production to explain their respective growth paths.  
This further leaves open the possibility for 
different explanations of growth in various countries
at different stages of their development, 
a central point of long-wave economic growth modeling.~\citep{Froling:2009vo, Jones:2001wn, Galor:2000up} 
**** The previous sentences sound like we'll show that some groups of countries
have similar charactersitics. 
That's what we thought we would find, 
but our new conclusion is that one must treat each country 
individually. 
We should consider revising the previous few sentences. --MKH ****

**** Suggest deleting the next two sentences:
Furthermore, there has been little empirical work assessing the predictive ability 
of various energy-augmented production functions for different types of economies. 
And, very few studies of the importance of energy in economic growth include 
rigorous statistical assessments of energy-related fitting parameters. 
End of deleted section --MKH ****

@@

The remainder of this paper is organized as follows. 
% Section~\ref{sec:lit_review} provides a necessarily brief overview of relevant literature.
Section~\ref{sec:Coordinates} discusses the coordinates of analysis for the present study. 
Section~\ref{sec:Sources_of_Data} details sources of data and calculations to 
generate time series for economic output ($y$) and factors of production ($k$, $l$, and $q$). 
The procedures for and results from parameter estimation for the various production functions 
are discussed in Section~\ref{sec:Parameter_Estimation}. 
Section~\ref{sec:Discussion} discusses implications of the parameter estimation 
results, and Section~\ref{sec:Conclusion} concludes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We find it beneficial to discuss the results through a series of questions, 
answers to which are illuminated by the techniques and data presented
in Section~\ref{sec:Parameter_Estimation} above.

**** We need to think about the order of the subsections below. --MKH ****

@@

For the Cobb-Douglas and CES production functions, 
augmentation by energy improves goodness of fit when $\gamma > 0$.%
  \footnote{
  The LINEX function is unhelpful in answering the question 
  ``Is it important to include energy in production functions?'' 
  because LINEX presupposes that energy is the only factor of production.
  }
First, we look at the results of the Cobb-Douglas functions 
in Figures~\ref{fig:CD_lambda_theta_graph} and~\ref{fig:CD_Ternary_Scatter_Graph}. 
Visually, $\gamma > 0$ is shown in the right column 
of Figure~\ref{fig:CD_Ternary_Scatter_Graph} 
as data points that are above the horizontal base of the ternary plots. 
We note that the list of countries for which $\gamma \gg 0$ includes 
Tanzania, South Africa, Saudi Arabia, and the United Kingdom. 
Especially for Tanzania, but also for Saudi Arabia and the United Kingdom, 
it appears certain that including energy improves goodness of fit: 
both the fit to historical data (crosshairs) 
and all resample data points (black dots) 
are above the base of the triangle in the energy columns of Figure~\ref{fig:CD_Ternary_Scatter_Graph}. 
However, for South Africa, significant uncertainty in $\gamma$, 
as shown by the large vertical spread of resample data points, 
leads us to exercise caution in concluding that 
augmenting the Cobb-Douglas production function by energy 
is helpful for improving the goodness of fit to historical data. 
A robust result 
($\gamma$ equal or close to zero and very little vertical spread of the resample points) 
indicates that including energy does not help to explain economic growth in Zambia,
while for United States, Japan, China, and especially Iran, 
it is less certain that augmenting with energy is unhelpful 
(larger vertical spread of resample points).

@@

Based on this analysis of the empirical data, 
we find no grounds to argue for augmenting the production function 
with energy for all economies. 
Our results support \citeapos{Stern:2011bq} conclusion that energy 
may or may not be required in production functions,
depending on its abundance or scarcity. 
One cannot know \emph{a-priori} the economies 
for which augmenting by energy will improve goodness of fit. 
Statistical testing is clearly warranted.

@@

Some developed economies (US) benefit little from including energy as a factor of production, 
while some others (UK) benefit in a minor way. 
For other developed economies (JP), the results of CD and CES models point in opposite directions.  
For NICs, the results are mixed; 
China does not convincingly benefit from energy-augmented production functions, 
and results are uncertain for South Africa (ZA) due to lack of precision in determining $\gamma$.  
In one OPEC country (SA), adding energy clearly helps to explain economic growth, 
while in another (IR), the benefit of energy augmentation is small and uncertain. 
For LDCs there is a either a strong indication that energy-augmented production functions 
are beneficial as an explanation for growth (TZ) or strong indication otherwise (ZM).

In summary, we observe that World Bank country categories (Table~\ref{tab:Economy_Examples}) 
do not predict whether energy augmentation will be beneficial. 
In fact, 
there is as much variation of the role of energy in economic growth 
within World Bank country categories
as between World Bank country categories. 
Thus, we conclude that the World Bank’s country categories are not helpful 
in explaining the role of energy in economic growth. 
Rather, a nuanced approach is necessary wherein the energy situation 
in each economy is assessed individually.

@@

The interpretation of elasticities of substitution 
among factors of production ($k$, $l$, and $e$) is challenging,
because a version of each elasticity of substitution appears three or four times 
among the various nests. 
For example, an elasticity of substitution between capital~($k$) and labor~($l$) 
appears in the following nests: 
($\underline{kl}$), 
($\underline{kl}$)$e$, 
($\underline{l}e$)$\underline{k}$, and 
($e\underline{k}$)$\underline{l}$. 
The following tables assist interpretation: 
C (c) indicates strong (weak) complementarity, 
S (s) indicates strong (weak) substitutability, and
$\sim\sim\sim$ ($\sim$) indicates large (small) uncertainty.

@@

Regarding Martin's question about space, 
I've seen other papers in \emph{Energy Economics} that are 51 printed journal pages. \citep{Saunders:2008aa}
I expect we are fine on length. --MKH ****

@@

**** For the time being, I've added two of these tables, US and UK.
We can add the remaining seven tables easily, either here or in an appendix. 
Data to fill the appendices is in the Word document on Dropbox. --MKH ****

@@

That so many estimates of elasticity of substitution exhibit large uncertainty, 
often spanning the range from perfect complements ($\sigma, \, \sigma_1 = 0$) 
to perfect substitutes ($\sigma, \, \sigma_1 = \infty$), 
is a troubling characteristic of the estimated CES coefficients 
that is rarely reported in the literature.
For some countries (IR, ZM), 
estimating any elasticity of substitution at all is problematic and fraught with peril. 
Note that this result is unlikely to be caused by poor-quality data alone. 
Presumably the best data in the world (US, UK, JP) exhibits large uncertainty. 
(See US and UK rows in Figures~\ref{fig:CD_Ternary_Scatter_Graph}  
and~\ref{fig:CES_sigma1_delta1_graph}--\ref{fig:CES_ternary_graph}.)

@@

Some of the uncertainty in parameter estimates may be explained by structural change in the economies. 
For example, 
**** Martin: can you select one of the highly uncertain economies and discuss it? 
Perhaps Iran or Zambia? ****. 
To account for the large uncertainty, 
it may be useful to perform piecewise analyses of smaller time periods. 
For the present study, doing piecewise analysis of, say, 
Iran **** use the economy that Martin discusses above **** 
would mean that the CES model is estimated on two 10-year periods. 
Considering that there are six fitted parameters in the CES model 
($\gamma$, $\lambda$, $\delta$, $\delta_1$, $\rho$, and $\rho_1$), 
ten years of historical data leaves precious few 
(only 4) degrees of freedom for the fitting process.

Based on the results and discussions above 
regarding the relative importance of and 
the elasticity of substitution among factors of production, 
we conclude that use of the CES model for energy analysis 
should be undertaken with utmost caution. 
Indeed, rigorous statistical testing on the relative importance 
of production factors and elasticities of substitution is highly recommended. 
Our results support \citeapos{Lecca:2011kr} conclusion 
for the need to econometrically specify production function relationships and 
to perform sensitivity analysis on nesting choices in economies 
to enhance the reliability of results.

@@

Our assessment: None of the production functions is perfect. 
Probably CD is best compromise among ease of estimating, good (but not best) fit to historical data, and
small uncertainties in fitted parameters.

If precise fitting to historical data is paramount (and you're unconcerned about uncertainties
in fitted coefficients), CES is the best.

If you need to represent rebound effects (not just backfire), 
CES must be employed, 
because the CES function was found by \citet{Saunders:2008aa} to be the only one of the production
functions studied here that is capable of representing rebound rebound in addition to backfire.
However, its ability to do so is tied to the values of the elasticity of substitution.
Unfortunately, we find that value of the elasticity of substitution is often highly uncertain.
Thus, whether the CES function is helpful for estimating both rebound and backfire effects
is questionable.

%++++++++++++++++++++++++++++++
\subsection{Implications of parameter uncertainty} 
\label{sec:implications}
%++++++++++++++++++++++++++++++

In addition to the usual parameter estimation process 
by linear coefficient determination (Cobb-Douglas and LinEx) 
and non-linear optimization algorithms (CES), 
the resampling/bootstrapping analysis performed herein 
provides insight into the question of how accurately parameters are known.

**** Move this paragraph elsewhere? --MKH ****
We chose the resampling approach instead of the more-common technique 
of estimating confidence intervals from standard errors for two reasons. 
(1)~Model parameters are highly constrained, and confidence intervals often violate the constraints. 
For example a result of $\alpha = 0.25 \pm 0.3$ is nonsensical, 
because $ \alpha \in [0, 1]$. 
(2)~True confidence intervals for constrained parameters are often asymmetric
(e.g., $\alpha = 0.2 + 0.1, -0.05$), 
but the standard error approach yields symmetric confidence intervals. 
The resampling approach that we employed both respects constraints 
and allows for asymmetric confidence intervals.%
  \footnote{
  We have chosen to present parameter uncertainty results visually, 
  both because the resampling technique lends itself to visual presentation and 
  because we believe visual comparison is more effective than tablefuls of numbers.
  }

A significant result to emerge from this work is that large parameter uncertainty 
does not necessarily lead to poor model fit to economic output. 
To choose one example of many, 
the value of $\gamma$ for the Cobb-Douglas model for Tanzania (0.59) 
shows a resampling range of 0.2--0.9. 
(See Figure~\ref{fig:CD_Ternary_Scatter_Graph}.) 
Yet, Cobb-Douglas models for Tanzania, with and without energy, both 
(a)~fit historical data very well 
(white line compared to black line in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}) 
and 
(b)~exhibit a very narrow range of resampled fits 
(thin gray band in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}). 
This means that a good fit to historical data does not indicate 
that model parameters have small uncertainty. 
In fact, similar to Tanzania, 
many economies exhibit both large uncertainty in model parameters and
very good fit to historical data. 
(See Figures~\ref{fig:CD_lambda_theta_graph} ~\ref{fig:CD_Ternary_Scatter_Graph} 
compared to Figure~\ref{fig:CD_GDP_Spaghetti_Graph}
for the Cobb-Douglas models and 
Figure~\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_sigma_delta_graph}
compared to Figure~\ref{fig:CES_Spaghetti_Graph} for CES models.) 
This result is consistent across almost all economies and models. 
Thus, we conclude that great care must be taken when interpreting estimated parameters: 
it is very easy to claim more certainty than is warranted. 
Where large uncertainty is observed, 
additional work is needed to contextualize the results, 
for example, 
sector-based analysis or ****. **** Martin, add more here? ****
